Bayesian exponential random graph modeling of KWL Burial network

### 2020-11-16 analysis/paper/paper.Rmd
edit assessments for MCMC and work on para for results

<!--burial study- network analysis-->
Burials analysis by archaeologists is an approach to understand past social structure through the study of the physical traces of mortuary practices. The material cultures and biological records of burial behaviors can represent the social status or identity of the deceased [@Binford1971; @Saxe1970]. Despite criticism that manipulation of burial rituals by the living can cause a disconnect between a person's status in life and their status represented by burial contexts [@Hodder1980; @Pearson1982], burials can still provide valuable information to infer past societies [@Chapman2003]. For example, mortuary practices, including burial forms, grave goods, and ritual behaviors, that structure the material configuration of burials can represent social relations between members of a community, allowing the inference of social structures [@Byrd1995; @Seikel2011]. Based on network science and graph theories, the recent development of social network analysis provides many new tools to visualize and analyze relationships for archaeological data [@Borgatti2009; @Brughmans2013]. Network science is the study of relational data where a phenomenon is conceptualized as a network through two steps, abstraction and representation, based on our specification in terms of network theories [@Brandes2013; @Collar2015]. That is, a network can be viewed as a patterned aggregation that includes individual elements (i.e. individual burials), pair-wise relationships (the dyads, for example burials with similar types and amounts of grave goods), and an overall structure showing global patterns represented in the data.

@@

<!-- archaeological specific case study -->
In this study, we use a pericolonial model to explore the changes in social inequality at Kiwulan, an Iron Age site in northeastern Taiwan, which covers the time before the European arrival, the presence of the Europeans, including the Spanish and the Dutch, in the 17th century, and finally the Chinese in the 19th century [@Chen2007]. We assume that the social changes at Kiwulan would be supported by chronological differences in the structure of burial networks. We examine burial networks before and after the foreign contacts, and test our hypothesis that increased social inequality can be observed in the network after the European arrival. We ask: (1) does the observed burial data after the start of European presence resemble the of network our hypothesized model? (2) did European colonial activities in 17th century Taiwan result in the emergence of social inequality in an Indigenous society in ways that can be detected by burials networks? (3) what are the major variables affecting or forming the higher degree of popularity (e.g. a few nodes have more relationships with other nodes) in the European contact period that might hint at social heterogeneity during this period? By answering these questions, this study helps to expand the use of burials in understanding the indirect effects of a colonial presence on Indigenous groups. This is important to understand the reactions of Kiwulan residents in the context of indirect impacts of European colonization. 
# Exponential random graph models in a Baysesian framework

@@

<!-- Network construction and specification -->
We used Bayesian inference on exponential random graph models (ERGMs) to quantify the relations among burials and test our hypothesis of social change using the R programming language [@Rlanguage2019] with the bergm package [@Caimo2014], which is built upon the statnet package for model specification and simulation procedures [@Hunter2008; @Morris2008; @Handcock2008]. Table \@ref(tab:ergm-terms-table) lists the configurations we used for the model specification for burials with the corresponding archaeological evidence. Every parameter in an ERGM has an associated algorithm for computing its value for a network [@Morris2008]. Based on our hypothesis, we model a network with increased social inequality that is represented by low transitivity and high popularity dyadic effects. We include nodal covariate effects for homophily, including age, gender, ritual activity, and the degree of wealth, to test whether nodes with similar attributes tend to form relationships presented by the network ties overall. We also take into account the physical distance between burials as an indicator of a kinship-based relation since the deceased from the same family tend to be buried nearby [@LiandWu2006]. Our model could reveal the emergence of social inequality via the presence of a few individuals as network centers, having more relations with others. This would be indicated by high popularity or degree values in the network statistics, with covariates showing the underlying mechanisms. According to our hypothesis, the burial evidence from after the European arrival will show higher popularity compared with the burial evidence from before European contact.

@@

<!-- add text about prior setting and anthropological assumptions of our prior values --> 
Normal distribution for the priors is typical in network analysis studies that assume networks to have low density and high transitivity, as are commonly found in the real world [@Caimo2017]. Thus, we specified the priors as low density for both network models. For covariate based on nodal attributes, such as age, gender, and wealth index, we set the priors to follow a normal distribution with mean 0 and standard deviation 5 (i.e. N(0, 5)) for both models. We set a negative covariate effect based on physical distance between burials (N(-1, 5)). To evaluate our anthropological assumption of increased social inequality over time, we incorporated different prior information for the network variables that are meaningful for examining social inequality, especially for gwesp and gwedegree. For example, we set the priors to higher transitivity (N(5, 7)), lower popularity (N(4, 7)), and higher covariate effect based on the ritual (N(2, 5)) for network before European contact to indicate less social inequality. On the contrary, we set the prior for the network after the European arrival to lower transitivity (N(4, 8)), and higher popularity (N(8, 8)) to illustrate an increased social inequality. This prior information derives from the theory about the social structure of horizontal hierarchy, which can be viewed as a spectrum that illustrates an increasing social inequality from a corporate model at one end toward a network mode at the other end [@Drennan2010; @Feinman2000]. The priors for density were also set differently with N(-17, 7) for earlier network and N(-20, 6) for later network. We should note that the mean priors and standard deviation priors indicate relative differences for comparison.

<!-- Goodness-of-fit assessment-->
The output from our two  models are assessed by goodness-of-fit (GOF) diagnostics in the Bayesian framework, where the observed network is compared with the set of networks simulated from the estimated posterior distributions of the parameters of the model [@Caimo2011; @Caimo2017]. Our Bayesian GOF summarized three distributions, including degree, minimum geodesic distance, and edgewise shared partner distributions. This provides a statistical approach to check how well the estimated posterior parameter distribution based on our hypotheses can reproduce networks similar to the general structural features of the observed network. We then compared the distribution of our observed networks, pre-European and post-European networks, with the distribution of our hypothesized model with higher popularity. We expect to see the network after the European presence has a better fit with the hypothesis, indicative of a structure with higher social differentiation.

@@

The pre-European network shows three subgroups that possessing the same types of burial goods: medium amount of carnelian bead on the right, medium amount of gold-foil beads on the left, and porcelains in the middle. Although burials with different amount of burial goods can be identified, there is no clear clusters demonstrating the sharing of prestige goods only for burials with rich goods. The post-European network also presented subgroups with more edges connecting burials.

>>>>

<!--burial study- network analysis-->
Burials analysis by archaeologists is an approach to understand past social structure through the study of the physical traces of mortuary practices. The material cultures and biological records of burial behaviors can represent the social ranking or identity of the deceased [@Binford1971; @Drennan2010; @Saxe1970]. Despite criticism that manipulation of burial rituals by the living can cause a disconnect between a person's status in life and their status represented by burial contexts [@Hodder1980; @Pearson1982], burials can still provide valuable information to infer past societies [@Chapman2003]. For example, mortuary practices, including burial forms, grave goods, and ritual behaviors, that structure the material configuration of burials can represent social relations between members of a community, allowing the inference of social structures [@Byrd1995; @Seikel2011]. Based on network science and graph theories, the recent development of social network analysis provides many new tools to visualize and analyze relationships for archaeological data [@Borgatti2009; @Brughmans2013]. Network science is the study of relational data where a phenomenon is conceptualized as a network through two steps, abstraction and representation, based on our specification in terms of network theories [@Brandes2013; @Collar2015]. That is, a network can be viewed as a patterned aggregation that includes individual elements (i.e. individual burials), pair-wise relationships (the dyads, for example burials with similar types and amounts of grave goods), and an overall structure showing global patterns represented in the data.

@@

<!-- archaeological specific case study -->
In this study, we use a pericolonial model to explore the changes in social inequality at Kiwulan, an Iron Age site in northeastern Taiwan, which covers the time before the European arrival, the presence of the Europeans, including the Spanish and the Dutch, in the 17th century, and finally the Chinese in the 19th century [@Chen2007]. We assume that the social changes at Kiwulan would be supported by chronological differences in the structure of burial networks. We examine burial networks before and after the foreign contacts, and test our hypothesis that increased social inequality can be observed in the network after the European arrival. We ask: (1) does the observed burial data after the start of European presence resemble the of network our hypothesized model? (2) did European colonial activities in 17th century Taiwan result in the emergence of social inequality in an Indigenous society in ways that can be detected by burials networks? (3) what are the major variables affecting or forming the higher degree of popularity (e.g. a few nodes have more relationships with other nodes) in the European contact period that might hint at social heterogeneity during this period? By answering these questions, this study helps to expand the use of burials in understanding the indirect effects of a colonial presence on Indigenous groups. This is important to understand the reactions of Kiwulan residents in the context of indirect impacts of European colonization.

# Exponential random graph models in a Baysesian framework

@@

<!-- Network construction and specification -->
We used Bayesian inference on exponential random graph models (ERGMs) to quantify the relations among burials and test our hypothesis of social change using the R programming language [@Rlanguage2019] with the bergm package [@Caimo2014], which is built upon the statnet package for model specification and simulation procedures [@Hunter2008; @Morris2008; @Handcock2008]. Table \@ref(tab:ergm-terms-table) lists the configurations we used for the model specification for burials with the corresponding archaeological evidence. Every parameter in an ERGM has an associated algorithm for computing the probability of observing relations in terms of grave goods between two burials. Based on our hypothesis, we model a network with increased social inequality that is represented by low transitivity and high popularity dyadic effects. We include burial-specific attributes as covariate effects for homophily, such as age, gender, ritual activity, and the degree of wealth, to test whether burials with similar attributes tend to form relationships. For example, age-homophily means people of the same age to have the same burial goods. We also take into account the physical distance between burials as an indicator of a kinship-based relation since the deceased from the same family tend to be buried nearby [@LiandWu2006]. Our model could reveal the emergence of social inequality via the presence of a few individuals as network centers, having more relations with others. This would be indicated by high popularity or degree values in the network statistics, with covariates to control the preferential tendency of formation of relationship. According to our hypothesis, the burial evidence from after the European arrival will show higher popularity compared with the burial evidence from before European contact.

@@

<!-- add text about prior setting and anthropological assumptions of our prior values --> 
Normal distribution for the priors is typical in network analysis studies that assume networks to have low density and high transitivity, as are commonly found in the real world [@Caimo2017]. Thus, we specified the prior of edge to low density for both network models. For covariates based on burial attributes, such as age, gender, and wealth index, we set the priors to follow a normal distribution with mean 0 and standard deviation 5 (i.e. N(0, 5)) for both models. For physical distance between burials, we set a negative covariate effect (N(-1, 5)) to infer the kinship-based proximity. To evaluate our anthropological assumption about increased social inequality over time, we incorporated different prior information for the network variables that are meaningful for social inequality, especially for gwesp and gwedegree representing transitivity and popularity. We set the priors to higher transitivity (N(5, 7)), lower popularity (N(4, 7)), and higher covariate effect based on the ritual (N(2, 5)) for network before European contact to indicate less social inequality. On the contrary, we set the prior for the network after the European arrival to lower transitivity (N(4, 8)), and higher popularity (N(8, 8)) to illustrate an increased social inequality. This prior information derives from the theory about horizontal hierarchy, which can be viewed as a spectrum that illustrates an increasing social inequality from a corporate mode at one end toward a network mode at the other end [@Drennan2010; @Feinman2000]. The priors for density were also set differently with N(-17, 7) for earlier network and N(-20, 6) for later network. It should be noted the idea of those settings is to reflect the relative differences between two networks, and the standard deviation was adjusted in terms of different networks to indicate uncertainty.

<!-- assessment of MCMC and Goodness-of-fit in Bayesian-->
To assess MCMC analysis, we first evaluate three diagnostics presented as visual summaries, including density plots, trace plots, and autocorrelation plots [@Hamra2013]. Those plots provide an informal way to diagnose model convergence by identifying any unexpected peaks or shapes in distribution. Then we access output from our two  models by goodness-of-fit (GOF) diagnostics in the Bayesian framework, where the observed network is compared with the set of networks simulated from the estimated posterior distributions of the parameters of each model [@Caimo2011; @Caimo2017]. Our Bayesian GOF summarized three distributions, including degree, minimum geodesic distance, and edgewise shared partner distributions. This provides a statistical approach to check how well the estimated posterior parameter distribution based on our hypotheses can reproduce networks similar to the general structural features of the observed network. We then compared the distribution of our observed networks, networks before and after the arrival of Europeans, with the distribution of our hypothesized models. We expect to see the models fit with our hypotheses, indicative of a increased social inequality after the foreign contact. In addition, the covariates can give some more clues for the underlying mechanisms for the formations of each network.

@@

We compared estimated statistics from the posterior distributions, networks before and after the European presence, to examine their differences in network structure. For the earlier network, the covariates having significant tendency to form relations between burials includes age-homophily and ritual-homophily. The age homophiligy shows negative effect that that indicates burials with the same age does not share similar prestige goods, but different age does instead. The positive value of ritual-homophily shows burial with riutal pottery also to form relations. Despite other covariates are positive such as sex and burial values, they does not show significant tendency to form relationships.
